---
title: "DATA608 | Knowledge and Visual Analytics"
author: "Gabriella Martinez"
date: "10/5/2022"
output: html_document
---

^[https://github.com/tidyverse/dplyr/issues/5214]
```{r}
# remove.packages("rlang")
# install.packages("rlang")
#install.packages("gghighlight")
```


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(leaflet)
library(tigris)
library(plotly)
library(shiny)
library(shinythemes)
library(rsconnect)
library(sf)
library(ggplot2)
library(gghighlight)
```

```{r}
rsconnect::setAccountInfo(name='gabbypaola',
			  token='6C8C97C7673C3F08BCAB0D6A30D000C1',
			  secret='HOYwFQGj+q4L7lMHPHLuS7n+39AMR48ewVLrRpbh')
```



```{r message=FALSE, warning=FALSE}
df <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv")

names(df) <- lapply(names(df), tolower)
```

```{r message=FALSE, warning=FALSE}
str(df)

unique(df$icd.chapter)

#df$ICD.Chapter <- as.factor(df$ICD.Chapter)
```

## Question 1
As a researcher, you frequently compare mortality rates from particular causes across different states. You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all states, from one cause (for example, Neoplasms (cancers)). Create a visualization that allows you to rank states by crude mortality rate for each cause of death.


Static Bar Graph of the values fixing `icd.chapter`, `year`, and the top 25 values to display using `top_n()`.
```{r}
df %>% filter(year == 2010 & 
                icd.chapter == 'Mental and behavioural disorders') %>%
                arrange(desc(crude.rate)) %>% 
                top_n(10) %>% 
    ggplot(aes(x=reorder(state, crude.rate), y=crude.rate))+
     geom_col(fill='#f69697') +
     coord_flip() +
     geom_text(aes(label=crude.rate), size = 2, hjust = -0.1)+
     labs(title = "Cause of Death: Mental and behavioural disorders", 
          subtitle = "Crude Death Rate Across the US in 2010",
          x= "State", 
          y = "Crude Rate") +
     theme_minimal()
```

For shiny bar plot to show *rank* but have option to not show all states at once: create an option to select top 10, 25, or all.


### Interactive Bar Graph

^[https://stackoverflow.com/questions/50814555/how-to-make-top-n-reactive-with-using-inputx-inside-the-function]
```{r}
#shiny app UI
ui <- fluidPage(
  titlePanel(
    "Examining Causes of Death Across the US from 1999-2010"
  ),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = 'year', 
        label = 'Year:', 
        choices = unique(df$year),
        selected = 2010), #sets the default display 2010 chosen to answer the question
      selectInput(
        inputId = 'cause', 
        label = 'Cause of Death:', 
        choices = unique(df$icd.chapter),
        selected = "Certain infectious and parasitic diseases"),
      selectInput(
        inputId = 'topN', #controls the displayed amount of values
        label = 'Choose Top States to display', 
        choices = c(10,25,51), #display 10, 25 or 51 at a time
        selected = 10), #default is 10
    ),
    mainPanel(
      plotOutput('plot1', height = "800px")
    )
  )
)

#shiny app server
server <- shinyServer(function(input, output, session){
  plot_data <- reactive({
    filter(df, year == input$year & icd.chapter == input$cause) %>% 
      top_n(n=as.integer(input$topN),wt=crude.rate)
  })
  
  #barplot
  output$plot1 <- renderPlot({
    ggplot(plot_data(), aes(x=reorder(state, crude.rate), y=crude.rate))+
      geom_col(fill='#f69697') + 
      coord_flip() +
      geom_text(aes(label=crude.rate), size = 4, hjust = -0.1)+
      labs(title=paste('Crude Mortality Rates for',input$cause),
           subtitle=paste('Top', input$topN, 'states'),
           x = "State", 
           y = "Crude Rate")+
      theme_minimal()
  }
  )
})
shinyApp(ui = ui, server = server)
rsconnect::deployApp()
```



### Map 

```{r message=FALSE, warning=FALSE}
#load in shapefile
states <- tigris::states(cb=T)

#change column name on shapefile for left join with death data
states <- states %>% rename("state" = "STUSPS")
```

^[https://stackoverflow.com/questions/73181511/shapefile-warnings-in-r]
```{r message=FALSE, warning=FALSE}
#st_crs(states)$epsg

#transform states shapefile to resolve error message
states <- st_transform(states, 4326)
```


```{r message=FALSE, warning=FALSE}
#merge
#states_merged <- df %>% left_join(states, by="State")

#need to use geo_join in order to get sf type dataframe for leaflet
states_merged <- geo_join(data_frame=df, spatial_data= states, 
                          by_df= "state", by_sp= "state", by = NULL, 
                          how = "inner")

#remove NA
states_merged <- subset(states_merged, !is.na(crude.rate))
```

# Create a Static Map keeping `year` and `icd.chapter` constant
```{r message=FALSE, warning=FALSE}
states_merged_filtered <- filter(states_merged, 
                                 year == 2010 & icd.chapter == 
                                   'Mental and behavioural disorders') %>% 
                                group_by(icd.chapter) %>% 
                                mutate(rank = rank(desc(crude.rate),
                                                   ties.method = "first"))
#%>%  arrange(desc(crude.rate)) %>% top_n(10, wt=crude.rate)

# #create color palette based on range of Crude.Rate
pal <- colorNumeric("Reds", domain=states_merged_filtered$crude.rate) 

# Setting up the pop up text
popup_sb <- with(states_merged_filtered,
                 paste(NAME, '<br>',
                       "Rank: ", rank, '<br>',
                       "Crude Rate: ", crude.rate ,'<br>',
                       "Deaths: ", deaths, '<br>',
                       "Population", population))
```

^[https://learn.r-journalism.com/en/mapping/census_maps/census-maps/]
```{r message=FALSE, warning=FALSE}
# Mapping it with the new tiles CartoDB.Positron
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(-98.483330, 38.712046, zoom = 4) %>%
  addPolygons(data = states_merged_filtered,
              fillColor = ~pal(states_merged_filtered$crude.rate),
              fillOpacity = 0.7,
              weight = 0.2,
              smoothFactor = 0.2,
              popup = ~popup_sb) %>%
  addLegend(pal = pal,
            values = states_merged_filtered$crude.rate,
            position = "bottomright",
            title = "Crude Rate")
```

^[https://stackoverflow.com/questions/57030746/shiny-leaflet-reactive-function-not-working]
This is what actually helped me fix it: ^[https://stackoverflow.com/questions/17002160/shiny-tutorial-error-in-r]
```{r message=FALSE, warning=FALSE}
#shiny app UI
ui <- fluidPage(
  titlePanel(
    "Examining Causes of Death Across the US from 1999-2010"
  ),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = 'year', 
        label = 'Year:', 
        choices = unique(states_merged$year),
        selected = 2010),
      selectInput(
        inputId = 'cause', 
        label = 'Cause of Death:', 
        choices = unique(states_merged$icd.chapter),
        selected = "Certain infectious and parasitic diseases"),
    ),
    mainPanel(
      leafletOutput(outputId = 'map', height = 800)
    )
  )
)

#shiny app server
server <- shinyServer(function(input, output, session){
  map_data <- reactive({
    filter(states_merged, year == input$year & icd.chapter == input$cause)%>% 
      group_by(icd.chapter) %>% 
      mutate(rank = rank(desc(crude.rate),ties.method = "first"))
  })
  
  output$map <- renderLeaflet({
    #create color palette based on range of Crude.Rate
    pal <- colorNumeric("Reds", domain=map_data()$crude.rate)
    #Setting up the pop up text
    popup_sb <- with(map_data(), 
                     paste(NAME, '<br>',
                           "Year: ", year, '<br>',
                           "Rank: ", rank, '<br>',
                           "Crude Rate: ", crude.rate ,'<br>',
                           "Deaths: ", deaths, '<br>',
                           "Population: ", population))
    #create map  
    map_data() %>% 
      leaflet() %>%
      addProviderTiles("CartoDB.Positron")%>%
      setView(-98.483330, 38.712046, zoom = 4) %>% 
      addPolygons( 
        fillColor = ~pal(map_data()$crude.rate), #reference map_data() 
        fillOpacity = 0.7, 
        weight = 0.2, 
        smoothFactor = 0.2, 
        popup = ~popup_sb) %>%
      addLegend(pal = pal, 
                values = map_data()$crude.rate, #reference map_data() 
                position = "bottomright", 
                title = "Deaths per 100,000 persons")
  }
  )
})
shinyApp(ui = ui, server = server)
deployApp()
```


## Question 2
Often you are asked whether particular states are improving their mortality rates (per cause) faster than, or slower than, the national average. Create a visualization that lets your clients see this for themselves for one cause of death at a time. Keep in mind that the national average should be weighted by the national population.  


```{r}
natl_avg <- df %>% 
  group_by(icd.chapter, year) %>% 
  mutate(crude.rate= round((sum(deaths) / sum(population)) * 100000, 1)) %>% 
  mutate(state= "US") %>% 
  select(icd.chapter, state, year, crude.rate) %>% 
  distinct(icd.chapter, year, .keep_all = TRUE) %>% 
  ungroup()

#natl_avg
```

```{r}
df_avg <- df %>% 
  select(icd.chapter,state,year,crude.rate) %>% 
  rbind(natl_avg)

# sanity check                
# df_avg %>% 
#   filter(state == "US")
```

Create line chart where only natl avg is highlighted
^[https://cran.r-project.org/web/packages/gghighlight/vignettes/gghighlight.html]
```{r}
#plots all states vs. US w/ cause as Mental and behavioural disorders
df_avg_plot <- df_avg %>% 
  filter(icd.chapter == 'Mental and behavioural disorders')

df_avg_plot %>% 
  ggplot(aes(year, crude.rate), color=state) +
  geom_line(aes(year, crude.rate, color = state), size=.7) +
  gghighlight(state=="US", label_params = list(size = 5),
              unhighlighted_params = list(colour="darkgray",
                                          alpha = 0.6,
                                          size=.6))+
  theme_minimal()
```

```{r}
#state==c("US","NJ")  | state == 'US'  ###%in% c('NJ','NY','MA')
#US will stay constant, the other state will be user input give user to select multiple

df_avg_plot <- df_avg %>% 
  filter(icd.chapter == 'Certain infectious and parasitic diseases' , 
         (state == 'US' | state %in% c('NJ','NY')) )

# df_avg_plot <- mutate(df_avg_plot, Label = ifelse(year == 2010, state, NA),
#             state = as.character(state))

df_avg_plot %>% 
  ggplot(aes(year, crude.rate), color=state) +
  geom_line(aes(year, crude.rate, color = state), size=.7)+
  theme_minimal() 
  
# geom_label(aes(label = Label), nudge_x = 0.35, size = 3) +
# gghighlight(state=="US", label_params = list(size = 5),
#               unhighlighted_params = list(colour="darkgray",
#                                           alpha = 0.6,
#                                           size=.6))
```


```{r}
df_avg_plot <- df_avg %>%
  filter(icd.chapter == 'Mental and behavioural disorders', 
         state == 'US' | state %in% c('NJ','NY','MA') )

p <- df_avg_plot %>%
  ggplot(aes(year, crude.rate), color=state) +
  geom_line(aes(year, crude.rate, color = state), size=.7) +
  theme_minimal() 

ggplotly(p) 
# %>% hide_legend()
```


```{r message=FALSE, warning=FALSE}
#shiny app UI
ui <- fluidPage(
  titlePanel(
    "Examining Causes of Death Across the US from 1999-2010"
  ),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        inputId = 'cause', 
        label = 'Cause of Death:', 
        choices = unique(df_avg$icd.chapter),
        selected = "Certain infectious and parasitic diseases"),
      selectInput(
        inputId = 'state', 
        label = 'State:', 
        multiple = TRUE,
        choices = unique(df_avg$state),
        selected='AL'),
    ),
    mainPanel(
      plotOutput(outputId = 'lineplot')
    )
  )
)

#shiny app server
server <- shinyServer(function(input, output, session){
  line_data <- reactive({
    df_avg %>% 
      filter(icd.chapter == input$cause, 
             (state == 'US' | state == input$state) )
  })
  
  output$lineplot <- renderPlot({
    
    #create plot
    ggplot(line_data(), aes(year, crude.rate)) +
      geom_line(aes(color = state), size=.7) +
      labs(x= "Year", y = "Crude Rate")+
      theme_minimal() 
    
    #ggplotly(p)
    #end plot
  }
  )
})
shinyApp(ui = ui, server = server)
deployApp()
```
