---
title: "DATA608 | Knowledge and Visual Analytics"
author: "Gabriella Martinez"
date: "12/7/2022"
output:
      html_document:
        toc: yes
        toc_float: yes
        theme: yeti
        highlight: kate
        font-family: "Arial"
        code_folding: show
---
# State Elected Political Parties and Unemployment Data

The purpose of this project was to explore the relationship between the political parties selected in the last two elections (including each year prior to the election) and unemployment rates and benefit amounts across the U.S..

# Packages
```{r message=FALSE, warning=FALSE}
library(readxl)
library(readr)
library(rvest)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(extrafont)
```

# Load, Clean, & Merge Data
## Unemployment Weekly Benefit Amounts

The unemployment maximum weekly benefit amounts were pulled from [Saving to Invest](www.savingtoinvest.com/maximum-weekly-unemployment-benefits-by-state/) in conjunction with the [Wayback Machine](https://web.archive.org/).  

Saving to Invest contains the list of unemployment insurance benefit amount/compensation rates by state and are pulled from each respective state website.  

The Federal-State Unemployment Compensation (UC) program is a partnership based upon federal law, but administered by state employees under state laws. Thus each state designs its own UC program within the guidelines of the federal requirements, which includes setting the benefit amount along with eligibility and disqualification provisions. There are significant differences between states so please visit the state unemployment for detailed rules and benefit calculation scenarios.  
  
The Internet Archive or Wayback Machine, is a 501(c)(3) non-profit that's building a digital library of Internet sites and other cultural artifacts in digital form. Like a paper library, it provides free access to researchers, historians, scholars, people with print disabilities, and the general public. It is their mission is to provide Universal Access to All Knowledge.

[2015](https://web.archive.org/web/20150913042755/https://savingtoinvest.com/maximum-weekly-unemployment-benefits-by-state/)  
[2016](https://web.archive.org/web/20161002140631/https://savingtoinvest.com/maximum-weekly-unemployment-benefits-by-state/)  
[2019](https://web.archive.org/web/20190903222914/https://savingtoinvest.com/maximum-weekly-unemployment-benefits-by-state/)  
[2020](https://web.archive.org/web/20201111210711/https://www.savingtoinvest.com/maximum-weekly-unemployment-benefits-by-state/)  

```{r message=FALSE, warning=FALSE}
#load unemployment benefits data

benefits15 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnempBenefits2015.csv")

benefits16 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnempBenefits2016.csv")

benefits19 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnempBenefits2019.csv")

benefits20 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnempBenefits2020.csv")

benefits15 <- benefits15 %>% select("State", "X2014.2015.Max.Weekly.Amt", "Year")
benefits16 <- benefits16 %>% select("State", "Max..Weekly.Benefit.Amount", "Year")
benefits19 <- benefits19 %>% select("State", "Max..Weekly.Benefit.Amount", "Year")
benefits20 <- benefits20 %>% select("State", "Max..Weekly.Benefit.Amount", "Year")

colnames(benefits15)[2] <- "Max..Weekly.Benefit.Amount"

benefits <- rbind(benefits15, benefits16, benefits19, benefits20)

colnames(benefits)[2] <- "MaxWeeklyBenefits"
```

## Unemployment Rates  

The unemployment rates were collected from the U.S. Bureau of Labor and Statistics (BLS). The data for years 2015 and 2016 were collected with the BLS site along with the Wayback Machine.

[2015](https://web.archive.org/web/20160309035527/http://www.bls.gov/lau/lastrk15.htm)  
[2016](https://web.archive.org/web/20170502003415/https://www.bls.gov/lau/lastrk16.htm)  
[2019](https://www.bls.gov/lau/lastrk19.htm)  
[2020](https://www.bls.gov/lau/lastrk20.htm)

```{r message=FALSE, warning=FALSE}
#load unemployment rate data

rates15 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnemploymentRates2015.csv")

rates16 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnemploymentRates2016.csv")

rates19 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnemploymentRates2019.csv")

rates20 <-read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/UnemploymentRates2020.csv")

rates <- rbind(rates15,rates16,rates19,rates20)
```

### Merge `rates` and `benefits` tables

```{r message=FALSE, warning=FALSE}
unemployment_df <- merge(x=benefits,y=rates,
                         by.x= c("State","Year"),
                         by.y= c("State","Year"))
```

## Election Results

The [Election Data](https://uselectionatlas.org/RESULTS/) was pulled from Dave Leip's Atlas of U.S. Presidential Elections. 

The 2012 election data was pulled in order to reflect the standing political party as well as how the states voted in 2015 prior to the 2016 election. In a similar fashion, a 2019 column was added to reflect the standing political party and the previously elected political party in each state prior to the 2020 election.


^[https://guides.loc.gov/election-statistics/presidential-elections]

```{r message=FALSE, warning=FALSE}
#load election data 

# needed for 2015 data
election12 <- read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/Election2012.csv")

election16 <- read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/Election2016.csv")

election20 <- read.csv("https://raw.githubusercontent.com/gabbypaola/DATA608/main/Final%20Project/Election2020.csv")
```

[2012 Battleground States](https://www.politico.com/2012-election/swing-state/)  

Swing states, also known as battleground states, are states that could “swing” to either Democratic or Republican candidates depending on the election. Because of their potential to be won by either candidate, political parties often spend a disproportionate amount of time and campaign resources on winning these states. ^[https://usafacts.org/articles/what-are-the-current-swing-states-and-how-have-they-changed-over-time/?utm_source=google&utm_medium=cpc&utm_campaign=ND-Elections&gclid=CjwKCAiAhKycBhAQEiwAgf19erVpBIcpduzLqacFnhYybgBM_hHDb9savQEByz143ymBmWdqnlDLzBoCHlgQAvD_BwE]  
Battleground states are states whose electoral history make them less predictable and typically feature comparatively close contests in terms of polling data and election results.^[https://ballotpedia.org/Presidential_battleground_states,_2016]

```{r message=FALSE, warning=FALSE}
# needed for 2015 data to show how the state voted in previous election 
election12 <- election12 %>% select("State", "Obama", "Romney", "Other")

election12$Obama = as.numeric(sub("%", "",election12$Obama))
election12$Romney = as.numeric(sub("%", "",election12$Romney))
election12$Other = as.numeric(sub("%", "",election12$Other))

election12$StateResult <- colnames(select(election12, "Obama", "Romney",
            "Other"))[max.col(select(election12, "Obama", "Romney","Other"))]

battleground_states12 =  data.frame("State"=c("Nevada","Colorado","Iowa",
                                              "Wisconsin", "Ohio","Virginia", 
                                              "North Carolina", "Florida",
                                              "New Hampshire"))

election12$SwingState <- ifelse(election12$State %in% 
                                  battleground_states12$State, "Yes", "No")

election12$StateElectedParty <- ifelse(election12$StateResult == "Obama", 
                                       "Democrat", "Republican")

#using 2015 as year to add standing president 
election12$Year = 2015

election12$ElectionResult = "Democrat"

election12 <- election12 %>% select("State", "Year", "StateResult", 
                                    "SwingState", "StateElectedParty", 
                                    "ElectionResult")

```

[2016 Battleground States](https://ballotpedia.org/Presidential_battleground_states,_2016)

```{r message=FALSE, warning=FALSE}
election16 <- election16 %>% select("State", "Clinton", "Trump", "Johnson",
                                    "Other")
#str(election16)

election16$Clinton = as.numeric(sub("%", "",election16$Clinton))
election16$Trump = as.numeric(sub("%", "",election16$Trump))
election16$Johnson = as.numeric(sub("%", "",election16$Johnson))
election16$Other = as.numeric(sub("%", "",election16$Other))

election16$StateResult <- colnames(select
                                   (election16,"Clinton", "Trump", "Johnson", 
                                     "Other"))[max.col(select(election16, 
                                                            "Clinton", "Trump", 
                                                            "Johnson","Other"))]

battleground_states16 = data.frame("State"=c("Arizona",
                                             "Nevada","Colorado","Iowa",
                                             "Wisconsin","Michigan","Ohio",
                                             "Pennsylvania","Virginia",
                                             "North Carolina", "Florida",
                                             "New Hampshire"))

election16$SwingState <- ifelse(election16$State %in% 
                                  battleground_states16$State, "Yes", "No")

election16$StateElectedParty <- ifelse(election16$StateResult == "Trump", 
                                       "Republican","Democrat")

election16$Year = 2016

election16$ElectionResult = "Republican"

election16 <- election16 %>% select("State", "Year", "StateResult", 
                                    "SwingState", "StateElectedParty", 
                                    "ElectionResult")

election16$State <- sub("D. C.", "District of Columbia",election16$State)

#2019 standing president
standing_pres19 <- election16

#changing year to 2019 to reflect president elect for that year
standing_pres19["Year"]=2019
```

[2020 Battleground States](https://ballotpedia.org/Presidential_battleground_states,_2020#Battleground_states)

```{r message=FALSE, warning=FALSE}
election20 <- election20 %>% select("State", "Biden", "Trump", "Other")

election20$Biden = as.numeric(sub("%", "",election20$Biden))
election20$Trump = as.numeric(sub("%", "",election20$Trump))
election20$Other = as.numeric(sub("%", "",election20$Other))

election20$StateResult <- colnames(
  select(election20, "Biden", "Trump", "Other"))[max.col(
    select(election20, "Biden", "Trump", "Other"))]

battleground_states20 = data.frame("State"=c("Arizona","Florida",
                                             "Georgia","Iowa","Michigan", 
                                             "Minnesota","Nevada",
                                             "New Hampshire","North Carolina",
                                             "Ohio","Pennsylvania","Texas",
                                             "Wisconsin"))

election20$SwingState <- ifelse(election20$State %in% 
                                      battleground_states20$State, "Yes", "No")

election20$StateElectedParty <- ifelse(election20$StateResult == "Trump", 
                                       "Republican","Democrat")

election20$Year = 2020

election20$ElectionResult = "Democrat"

election20 <- election20 %>%  select("State", "Year", "StateResult", 
                                     "SwingState", "StateElectedParty", 
                                     "ElectionResult")

election20$State <- sub("D. C.", "District of Columbia",election20$State)
```

### Merge `unemployment_df` with `election`

```{r message=FALSE, warning=FALSE}
#merge all election data
election <- rbind(election12, election16, standing_pres19, election20)

#merge `unemployment_df` with `election`
unemp_elec_df <- unemployment_df %>% full_join(election,
                           by= c("State","Year"))

colnames(unemp_elec_df)[4] <- "UnemploymentRate"

colnames(unemp_elec_df)[5] <- "UnemploymentRank"
```

### Add Census Regions and merge

[Census Regions and Divisions of the United States](https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf)

```{r message=FALSE, warning=FALSE}
regions <- read.csv("https://raw.githubusercontent.com/cphalpert/census-regions/master/us%20census%20bureau%20regions%20and%20divisions.csv")

regions <- regions %>% select("State","State.Code","Region")

unemp_elec_df <- unemp_elec_df %>% full_join(regions, by="State")

#split data up into pre-election year (previous election results)

pre_election_df <- unemp_elec_df %>% filter(Year %in% c(2015,2019))

election_result_df <- unemp_elec_df %>% filter(Year %in% c(2016,2020))

```

# Static Visualizations

## Rates vs. Benefit Amounts


^[https://rdrr.io/github/joon-e/allespaletti/src/R/politics.R]  
^[https://flowingdata.com/2017/06/12/politics-of-religious-leaders/]
```{r message=FALSE, warning=FALSE, fig.align = 'center'}
unemp_elec_df %>% filter(Year==2015) %>% 
  ggplot(aes(x=UnemploymentRate, y=MaxWeeklyBenefits, color=StateElectedParty)) +
  geom_point(size= 2) +
  geom_smooth(method="lm",se = FALSE)+
  scale_color_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027")) +
  # scale_color_manual(values= c("Democrat" = "#000E89",
  #                              "Republican" = "#C91602"))+
  # scale_color_manual(values= c("Democrat" = "#3885D3",
  #                              "Republican" = "#E02E44"))+
  guides(color = guide_legend(title = "Elected Party")) +
  labs(title="Unemployment Rate vs. Unemployment Weekly Benefit Amounts",
       subtitle= "2015",
       x= "Unemployment Rate", 
       y = "Weekly Benefit Amount")+
  theme(plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        panel.grid.minor.y = element_blank())
```

## Unemployment Benefit Amounts

```{r message=FALSE, warning=FALSE}
unemp_elec_df %>% filter(Year==2015) %>% 
  group_by(StateElectedParty, Region) %>% 
  summarise(AvgBenefits = mean(MaxWeeklyBenefits))
```

^[https://www.learnbyexample.org/r-bar-plot-ggplot2/#:~:text=For%20grouped%20bars%2C%20there%20is,to%20be%20larger%20than%20width.]
```{r message=FALSE, warning=FALSE, fig.align = 'center'}
unemp_elec_df %>% filter(Year==2015) %>% 
  group_by(StateElectedParty, Region) %>% 
  summarise(AvgBenefits = round(mean(MaxWeeklyBenefits),2)) %>% 
  ggplot(aes(fill=StateElectedParty, y= AvgBenefits, 
             x=reorder(Region, AvgBenefits)))+
  geom_col(stat="identity", width=0.7,
           position=position_dodge(width = 0.8))+ #,  alpha=0.7 position = "dodge" 
  # geom_text(
  #   aes(label = AvgBenefits),
  #   colour = "black", size = 3,
  #   vjust = 0.5, hjust=-0.1 , position = position_dodge(.9)
  # )+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
   # scale_fill_manual(values= c("Democrat" = "#000E89",
   #                             "Republican" = "#C91602"))+
  # scale_fill_manual(values= c("Democrat" = "#3885D3",
  #                              "Republican" = "#E02E44"))+
  guides(fill=guide_legend(title="Elected Party"))+
  coord_flip()+
  labs(title="Regional Previous Election Results and Average Unemployment Benefits",
       subtitle= "2015",
       y= "Average Weekly Benefits", 
       x = "Region")+
  theme(plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```


```{r fig.height=8, fig.width=10, message=FALSE, warning=FALSE, fig.align = 'center'}


unemp_elec_df %>% filter(Year==2015) %>% 
  group_by(StateElectedParty, Region) %>%
  mutate(avg.benefit = mean(MaxWeeklyBenefits)) %>% 
  ggplot(aes(x=reorder(State,MaxWeeklyBenefits),y=MaxWeeklyBenefits, 
             color=StateElectedParty, 
             group=StateElectedParty, 
             fill=StateElectedParty))+
  facet_wrap(~Region, scales='free_y')+ 
  geom_col(width= 0.5, colour=NA) +
  #   geom_text(
  #   aes(label = MaxWeeklyBenefits),
  #   colour = "white", size = 3,
  #   vjust = 0.5, hjust=1.2 , position = position_dodge(.5)
  # )+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
  # scale_fill_manual(values= c("Democrat" = "#000E89",
  #                              "Republican" = "#C91602"))+
  theme(axis.line=element_line())+
  labs(y= "Maximum Weekly Benefits", x = "State")+
  guides(fill=guide_legend(title="Elected Party"))+
  coord_flip()+
  labs(title="Previous Election Results and Average Unemployment Benefits",
       subtitle= "2015",
       y= "Average Weekly Benefits", 
       x = "Region")+
  theme(plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) 

```


```{r fig.height=8, fig.width=10, message=FALSE, warning=FALSE, fig.align = 'center'}


unemp_elec_df %>% filter(Year==2015, SwingState=="Yes") %>% 
  group_by(StateElectedParty, Region) %>%
  mutate(avg.benefit = mean(MaxWeeklyBenefits)) %>% 
  ggplot(aes(x=reorder(State,MaxWeeklyBenefits),y=MaxWeeklyBenefits, 
             color=StateElectedParty, 
             group=StateElectedParty, 
             fill=StateElectedParty))+
  facet_wrap(~Region, scales='free_y')+ 
  geom_col(width= 0.5, colour=NA) +
  #   geom_text(
  #   aes(label = MaxWeeklyBenefits),
  #   colour = "white", size = 3,
  #   vjust = 0.5, hjust=1.2 , position = position_dodge(.5)
  # )+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
  # scale_fill_manual(values= c("Democrat" = "#000E89",
  #                              "Republican" = "#C91602"))+
  theme(plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())+
  labs(title="Previous Election Results and Maximum Unemployment Benefits- Swing States",
       subtitle= "2015",
       y= "Maximum Weekly Benefits", 
       x = "State")+
  guides(fill=guide_legend(title="Elected Party"))+
  coord_flip()
```


## Unemployment Rates

```{r message=FALSE, warning=FALSE, fig.align = 'center'}


unemp_elec_df %>% filter(Year==2015) %>% 
  group_by(StateElectedParty, Region) %>% 
  summarise(AvgUnemp = round(mean(UnemploymentRate),2)) %>% 
  ggplot(aes(fill=StateElectedParty, y= AvgUnemp, 
             x=reorder(Region, AvgUnemp)))+
  geom_col(stat="identity", width=0.7,
           position=position_dodge(width = 0.8))+ #,  alpha=0.7 position = "dodge" 
  # geom_text(
  #   aes(label = AvgUnemp),
  #   colour = "black", size = 3,
  #   vjust = 0.5, hjust=-0.1 , position = position_dodge(.9))+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
   # scale_fill_manual(values= c("Democrat" = "#000E89",
   #                             "Republican" = "#C91602"))+
  # scale_fill_manual(values= c("Democrat" = "#3885D3",
  #                              "Republican" = "#E02E44"))+
  guides(fill=guide_legend(title="Elected Party"))+
  labs(title="Previous Election Results and Unemployment Rates",
       subtitle= "2015",
       y= "Average Unemployment Rate", 
       x = "Region")+
  coord_flip()+ 
  theme(plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        #panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
  
```


```{r fig.height=8, fig.width=10, message=FALSE, warning=FALSE, fig.align = 'center'}
unemp_elec_df %>% filter(Year==2015) %>% 
  group_by(StateElectedParty, Region) %>%
  mutate(avg.unemp = mean(UnemploymentRate)) %>% 
  ggplot(aes(x=reorder(State,UnemploymentRate),y=UnemploymentRate, 
             color=StateElectedParty, 
             group=StateElectedParty, 
             fill=StateElectedParty))+
  facet_wrap(~Region, scales='free_y')+ 
  geom_col(width= 0.5, colour=NA) +
  #   geom_text(
  #   aes(label = UnemploymentRate),
  #   colour = "white", size = 3,
  #   vjust = 0.5, hjust=1.2 , position = position_dodge(.5)
  # )+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
  # scale_fill_manual(values= c("Democrat" = "#000E89",
  #                              "Republican" = "#C91602"))+
  theme(axis.line=element_line(),
        plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        #panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())+
  labs(title="Previous Election Results and Unemployment Rates",
       subtitle= "2015",
       y= "Average Unemployment Rate", 
       x = "State")+
  guides(fill=guide_legend(title="Elected Party"))+
  coord_flip()
```




```{r fig.height=8, fig.width=10, message=FALSE, warning=FALSE, fig.align = 'center'}
unemp_elec_df %>% filter(Year==2015, SwingState=="Yes") %>% 
  group_by(StateElectedParty, Region) %>%
  mutate(avg.unemp = mean(UnemploymentRate)) %>% 
  ggplot(aes(x=reorder(State,UnemploymentRate),y=UnemploymentRate, 
             color=StateElectedParty, 
             group=StateElectedParty, 
             fill=StateElectedParty))+
  facet_wrap(~Region, scales='free_y')+ 
  geom_col(width= 0.5, colour=NA) +
  # geom_text(
  #   aes(label = UnemploymentRate),
  #   colour = "white", size = 3,
  #   vjust = 0.5, hjust=1.2 , position = position_dodge(.5)
  # )+
  scale_fill_manual(values= c("Democrat" = "#4575b4",
                               "Republican" = "#d73027"))+
  # scale_fill_manual(values= c("Democrat" = "#000E89",
  #                              "Republican" = "#C91602"))+
  theme(axis.line=element_line(),
        plot.title = element_text(size = 12, color = "#000500"),
        plot.subtitle = element_text(size = 9, color = "#484e4c"),
        # panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())+
  labs(title="Previous Election Results and Unemployment Rates- Swing States",
       subtitle= "2015",
       y= "Average Unemployment Rate", 
       x = "State")+
  guides(fill=guide_legend(title="Elected Party"))+
  coord_flip()
```


<!------- Below is for removing excessive space in Rmarkdown | HTML formatting -------->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
